Parameter-less Population Pyramid

### 2013-12-17 Parameterless Population Pyramid/main.tex
Add it all, who cares.

\newcommand{\includegraphicswide}[1]
{\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{#1}}

@@

\title{Parameterless Population Pyramid}
\subtitle{[Genetic Algorithms Track]}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\maketitle
\begin{abstract}
TODO
\end{abstract}

@@

\keywords{Linkage Learning, Local Search, TODO}

\section{Introduction}
TODO Parameterless optimization, model building, etc.

TODO Talk about ALPS~\cite{hornby:2006:alps}.

TODO Talk about LO-LIMD~\cite{posik:2011:parameterless} and how it was parameter-less,
but is incapable of handling overlapping linkages.

\section{The P3 Algorithm}
\label{sec-p3}
Before going into detail about how each aspect of P3 works, consider the example
evolution presented in Figure~\ref{fig-p3}.  At the start of optimization there
is no known problem information, so solution \#1 is randomly generated.  This
is then brought to a local optima using a hill climber (Section~\ref{sec-hillclimber})
creating solution \#2, which is added to the pyramid (Section~\ref{sec-pyramid}) level
$P_0$.  As there is no further exploitable information, \#3 is randomly generated
and optimized to create \#4, which is added to $P_0$.  \#2 and \#4 are used to
define a crossover (Section~\ref{sec-crossover}), which improves \#4 to create
\#5. As \#5 used information from $P_0$ is it added to the next level of the pyramid, $P_1$.
Again P3 lacks information to continue, so \#6 is randomly generated and improved into
\#7.  Unlike \#2 and \#4, \#7 is not added in this example because it happens to be identical to
a solution already in the pyramid.  Even though \#7 is not unique, it still crosses with
$P_0$, which does make a unique improvement \#8 that is added to $P_1$ and mixed with that
level to create \#9.  As before, the process creates \#10, \#11, and \#12 using
random generation, hill climbing, and crossover, respectively.  In this example \#13
is not added to $P_2$ because it not a strict improvement in fitness over \#12.  There
is not enough information in $P_2$ to improve \#13 further, so the process would begin
yet again with a random solution.

@@

\subsection{Pyramid}
\label{sec-pyramid}
Unlike many EC techniques, P3 does not have a single population of solutions.
Instead to maintains a pyramid like structure, such that each level of the pyramid
has a population of solutions, with each level representing something akin to different
generations of evolution.  The exact relationship between the levels is described
in Section~\ref{sec-alltogether}.

@@

Starting from the leaves, such that each loci appears exactly once, the binary tree
can be constructed using an agglomerative method.  Each iteration determine the
pairwise linkage between each cluster of loci using Equation~\ref{eq-distance},
and merge the two closest into a new cluster.  The process ends when there is only
a single cluster containing all loci.  As our clustering algorithm provides the proper
criteria this algorithm requires $O(N^2)$ total time to correctly construct the linkage tree
from the frequency table~\cite{gronau:2007:upgma}.

Not all of the clusters in the linkage tree are worth using during crossover.  The most
obvious is a cluster containing all the loci.  If we performed a crossover such that all
of the loci must remain together, no new individual would be created.  We further remove
clusters that are completely contained within some other cluster, such that if
$D(C_i,C_j) = 0$, $C_i$ and $C_j$ are not used when performing crossover.  $D(C_i,C_j)$ only
returns $0$ when the loci in $C_i$ perfectly predict the bits in $C_j$ and vice versa,
and from this we conclude that the settings for those two loci should never be disrupted.
When used on particularly small populations this has the potential to greatly reduce the number
of clusters used during crossover.  For instance, if $|P_i| = 1$ no clusters exist, and if $|P_i| = 2$
at most two clusters exist: one where the solutions agree and one where they disagree.

Given the useful set of clusters, we can now define how they are used.  P3 uses
crossover to recombine known information in a level of the pyramid with a candidate
solution.  For each cluster, do the following.  Select a random donor from the population.
Copy the donor's gene values for all loci in the cluster into the candidate solution.
Evaluate the candidate.  If the fitness of the candidate has been reduced, revert the candidate's
gene values.  When choosing the donor P3 ensures there is at least one gene value
different between the donor and the candidate solution.  This can potentially require
up to $O(|P_i|)$, but is on average more efficient than allowing donations that
do not change any gene values.

To complete a single crossover event, all of the useful clusters are tested, each
requiring at most a single evaluation.  Clusters are tested in smallest first order,
with clusters of equal size ordered randomly.  This ensures smaller building blocks
are optimized before large changes are made to the genome, which helps preserve
diversity.  As cluster building and application ordering are not dependent on
the position of loci in the genome, there is no positional bias in the search points
evaluated.

@@

Similarly, by starting from a random position for each new solution, P3 progressively reduces
the chance of spurious linkage formation due to initialization.  Furthermore, these random
restarts provide exploration to balance out the hill climber and crossover's extreme
exploitation.

@@

@@

If P3's quality relies solely on starting at random points and applying hill climber,
this comparison algorithm should perform even better than P3.  Similar to P3 this
algorithm is parameterless, makes no assumptions about gene ordering, and will
find the global optima given a sufficient (potentially exponential) number of evaluations.

@@

@@

In choosing test problems, we decided to use only benchmarks with knowable global
optima.  This allows us to measure how many evaulations each method requires to converge to the
global optima, as opposed to just the highest fitness reached.  This is advantageous
as three of the four algorithms have no hard termination condition.  The remaining
algorithm, LTGA, uses a population parameter which can be increased to slow
convergence with the potential increase in fitness reachable before convergence.
Therefore it is difficult to say how and when to declare any of the techniques
converged unless the global optimum is found.
While in practice it is often more important to find a good solution quickly
than to find the optimum after an exceedingly long time, using this as a measure
for algorithm quality can lead to algorithms which prematurely converge.

@@

The classic example for this kind of test problem is the Deceptive Trap problem.
For this benchmark the genome is broken up into non-overlapping problems of $k$
genes, with the fitness of a single ``trap'' given by Equation~\ref{eq-deceptivetrap}.
\begin{equation}
   trap(t) = \left\{
     \begin{array}{rl}
       k-1-t, &  t<k\\
       k,   &  t = k
     \end{array}
   \right.
  \label{eq-deceptivetrap}
\end{equation}
In this equation $t$ represents the number of bits set in the trap.  For each trap
there is one local optima (all bits set to 0) and one global optima (all bits set to 1).
This problem requires exponential time to solve unless there is some method to effectively
move complete traps between solutions, making this an excellent crossover test problem.

However, when using a hill climber and linkage learning, the Deceptive Trap problem
may be trivially solved~\cite{goldman:2012:ltga}.  As such the Deceptive Step Trap
problem was defined, given in Equation~\ref{eq-deceptivesteptrap}.
\begin{equation}
   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
  \label{eq-deceptivesteptrap}
\end{equation}
This problem leverages the original Deceptive Trap problem definition, with the
addition of fitness plateaus of configurable size $s$.  This creates an enormous
number of local optima and makes detection of linkages much more challenging.

@@

NK landscapes define a class of problems such that the fitness of a single gene's
value is dependent on $k$ other genes in the genome.  Finding the solution which
maximized fitness on an arbitrary NK landscape requires exponential time, and
are therefore a good candidate for heuristic techniques.  Nearest Neighbor NK
landscapes are a subset of NK landscapes such that each gene is dependent on the $k$
genes following it in the genome.  The advantage of this subset is that they can be
solved in polynomial time~\cite{wright:2000:solvingnk}.  For our experiments we
used wrapping neighborhoods, such that genes near the end of the genome depend on
genes at the start of the genome.

@@

Our final problem class is a subset of the Maximum Satisfiability (MAXSAT) problem.
This problem consists of a series of disjunctive clauses, such that each clause contains
three problem variables, some of which are negated. While
MAXSAT requires exponential time to solve, we have devised a method of generating
instances that ensures each has at least one solution that satisfies all clauses.
First, a random target solution is generated.  For each clause, set the sign of each
variable in the clause such that that clause will be satisfied by the target solution.
To ensure there is no bias to variable signs in the generated problem there is a $\frac{1}{6}$ probability
all clause signs match the target, a $\frac{1}{6}$ probability two of the three signs match,
and a $\frac{4}{6}$ probability only one sign matches.  By setting the clause signs
in this way we ensure the target solution is always able to satisfy all clauses,
meaning all problems have a known best fitness.  Note that its possible there to be
other solutions that also satisfy all clauses.

\section{Experimental Results}
\begin{figure*}
  \centering
  \includegraphicswide{WithRast}
  \caption{Test problems}
  \label{fig-results}
\end{figure*}

@@

  \end{tabular}
	\caption{Results}
	\label{table-results}
\end{table}

TODO Cover all 7 problems as a group, focusing on the two ``types'' of problems.
Make it explicit that you outperform by a constant factor on the static problems
and by an order of complexity on generated problems.

TODO Mention 1.5 trillion evaluations and 150,000 runs (average 10 million evaluations per run).

TODO Compare LTGA results with previously published results as best as possible.

TODO Mention source code is available\footnote{\url{https://name.of.site.removed/anonymous}}.

\section{Promising Theory}
TODO Easy proof on unimodal (no plateaus) and all functions of unitation due to HC.
Sketch of how to prove runtime on Deceptive Trap, HIFF, suggestions toward all
possible non-overlapping functions.

Due to P3's persistent integration of random solutions, it always has a non zero
probability of reaching the global optimum.  Because it meets this prerequisite
(unlike LTGA), it is likely possible to derive expected running times for P3 on
a number of problems.  While we do not present a rigorous solution to any particular
problem here, we wish to illuminate potential avenues for future proofs.

The traditional One Max class, as well as all functions of unitation, will be
trivially solved by the hill climber in at most $N+1$ evaluations.  These come
from the initial evaluation of the random solution, plus the $N$ evaluations required
to test each gene for improvement.  While more difficult to bound, any unimodal
search space without fitness plateaus will be solved by a single application
of the hill climber.  Notably we expect a bound on Leading Ones of $O(N^2)$ time.

Yet unlike many theory based algorithms, we believe P3 can achieve bounds on more
complex and interesting landscapes.  Consider a sketch for a proof on the Deceptive
Trap problem.  Each restart will require at most $2N$ evaluations to evaluate the
random individual and optimize it to a local optima using the hill climber.  In
$P_0$ all solutions will either have all 1s or all 0s.  As a result $D(C_i, C_j) = 0$
if all $C_i$ and $C_j$ come from the same trap.  Therefore all crossover clusters
will move only complete traps between solutions, and once there is sufficient
diversity in the population there will be a crossover for each individual trap.
When performing a crossover of a single trap there can only be a single alternative
value, meaning if the candidate solution has all 0s the donor must have all 1s, or vice
versa.  Therefore if $P_i$ contains the optimum solution for a trap in at least one
individual and has a crossover that moves that trap, all solutions produced by crossing
$P_i$ with a candidate solution will have the optimized value of that trap.  Extending to
all traps we can say that once any $P_i$ has at least one solution with the optimized
version of a trap for all traps the very next crossover with $P_i$ will produce the global optimum.
The only remaining portion of the proof is to determine how many evaluations are required to reach this point,
which is relatively trivial.

Using loose requirements, this can likely be extended to any arbitrary fully decomposable
problem. Assuming that for sufficiently large population sizes the entropy between
bits in a single subproblem is lower than the entropy of bits between subproblems,
P3 will construct a cluster capable of moving exactly one subproblem's gene values
during crossover.  As crossover cannot decrease the fitness of a solution each
crossover will have a non zero probability of improving subproblem solution quality.
The probability of low fitness local optima for a subproblem reaching high levels
is therefore low, which increases the probability of donors only having high quality
solutions to subproblems.  Combined this is expected to create a runtime potentially
exponential in subproblem size, but polynomial in problem size.

>>>>

\newcommand{\includegraphicswide}[1]
{\includegraphics[width=.9\textwidth,height=\textheight,keepaspectratio]{#1}}

@@

\title{Parameter-less Population Pyramid}
\subtitle{[Genetic Algorithms Track]}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\maketitle
\begin{abstract}
%Maximum 200 Words
A major limiting factor to many evolutionary techniques is their need to be reconfigured
to each new problem.  While some previous methods been able to remove or reduce
the need for parameter tuning, many do so by trading peak efficiency for general applicability.
The Parameter-less Population Pyramid (P3) strikes a balance between continuous integration
of diversity with extremely exploitative elitist operators, allowing it to solve easy
problems quickly and hard problems eventually.  When compared with three state of the art
optimization techniques across seven different problem classes P3 always finds the optimum
at least a constant factor \emph{\textbf{faster}} than the rest.  Even when optimally
tuning the comparisons, P3 has a lower growth complexity when test problems come
from a random distribution.  
Beyond empirical results, we suggest avenues for theoretical analysis of P3's
expected runtime.  In general, while these are only the first results for P3,
we feel they are indicative wide applicability.
\end{abstract}

@@

\keywords{Linkage Learning, Local Search, Parameter-less}

\section{Introduction}
Providing a good solution in reasonable time to a large set of real world black box problems
is a fundamental goal of evolutionary computation.  Achieving this goal requires an
optimizer to make assumptions about the nature of the problem is is solving.
%TODO NFL Citation?
Often built into the design of a genetic algorithm (GA) are evolutionary parameters,
which allow the user to tune the optimizer's assumptions to a specific problem.  These
parameters can allow a GA to efficiently search a multitude of landscapes.  However,
to do so the user must first determine how to correctly tune the parameters, which is
a complex and time consuming process unto itself.

The problem of parameter configuration has given rise to a number of attempts at
parameter-less optimization.  While many parameters (when using reasonable settings) are unlikely to effect the order
of complexity of an algorithm, population size can.  More specifically, sufficiently small
population sizes can cause exponential search times, with arbitrarily large population sizes
similarly stalling search.  Even more critically, setting the population size correctly relies
on difficult to discern features of each problem. In an early example, a crossover based GA was devised
which used concurrent racing populations to remove all user specified parameters~\cite{harik:1999:parameterlessga}.
This method of population racing has also been paired with more advanced evolutionary system,
such as Hierarchical Bayesian Optimization~\cite{pelikan:2004:parameterlesshboa}. The claim was that while not optimal,
this method ensured the GA would take no more than a logarithmic increase in evaluations.
Furthermore, doing so would ensure that these optimizers could be generally applicable
without problem specific tuning.  Experimental evidence supports this theory,
with some evidence that in general the bound is closer to a constant factor worse
than optimal.

In this work we shall introduce the Parameter-less Population Pyramid (P3).  This iterative
method for constructing a collection of populations requires no user parameters.
It leverages entropy based linkage detection to learn how to efficiently mix
solutions without problem specific information beyond the evaluation function.  Perhaps
most importantly, unlike other parameter-less techniques it experimentally appears to be at least
a constant factor \emph{\textbf{improvement}} over comparable optimally configured
optimization methods.

\section{The P3 Algorithm}
\label{sec-p3}
Before going into detail about how each aspect of P3 works, consider the example
evolution presented in Figure~\ref{fig-p3}.  At the start of optimization there
is no known problem information, so solution \#1 is randomly generated.  This
is then brought to a local optima using a hill climber (Section~\ref{sec-hillclimber})
creating solution \#2, which is added to the pyramid (Section~\ref{sec-pyramid}) level
$P_0$.  As there is no further exploitable information, \#3 is randomly generated
and optimized to create \#4, which is added to $P_0$.  \#2 and \#4 are used to
define a crossover (Section~\ref{sec-crossover}), which improves \#4 to create
\#5. As \#5 used information from $P_0$ it is added to the next level of the pyramid, $P_1$.
Again P3 lacks information to continue, so \#6 is randomly generated and improved into
\#7.  Unlike \#2 and \#4, \#7 is not added in this example because it happens to be identical to
a solution already in the pyramid.  Even though \#7 is not unique, it still crosses with
$P_0$, which does make a unique improvement \#8 that is added to $P_1$ and mixed with that
level to create \#9.  As before, the process creates \#10, \#11, and \#12 using
random generation, hill climbing, and crossover, respectively.  In this example \#13
is not added to $P_2$ because it not a strict improvement in fitness over \#12.  There
is not enough information in $P_2$ to improve \#13 further, so the process would begin
yet again with a random solution.

@@

\subsection{Pyramid}
\label{sec-pyramid}
Unlike many EC techniques, P3 does not have a single population of solutions.
Instead to maintains a pyramid like structure, such that each level of the pyramid
has a population of solutions, with each level representing something akin to different
generations of evolution.  More specifically, a solution's level is one higher than
its highest parent, similar to the layering done in~\cite{hornby:2006:alps}.

@@

Starting from the leaves, such that each loci appears exactly once, the binary tree
can be constructed using an agglomerative method.  Each iteration determine the
pairwise linkage between each cluster of loci using Equation~\ref{eq-distance},
and merge the two closest into a new cluster.  The process ends when there is only
a single cluster containing all loci.  As our clustering algorithm is both convex and
satisfies the swapping lemma~\cite{gronau:2007:upgma},
this algorithm requires $O(N^2)$ total time to correctly construct the linkage tree
from the frequency table.

Not all of the clusters in the linkage tree are worth using during crossover.  The most
obvious is a cluster containing all the loci.  If we performed a crossover such that all
of the loci must remain together, no new individual would be created.  We further remove
clusters that are completely contained within some other cluster, such that if
$D(C_i,C_j) = 0$, $C_i$ and $C_j$ are not used when performing crossover.  $D(C_i,C_j)$ only
returns $0$ when the loci in $C_i$ perfectly predict the bits in $C_j$ and vice versa,
and from this we conclude that the settings for those two loci should never be disrupted.
When used on particularly small populations this has the potential to greatly reduce the number
of clusters used during crossover.  For instance, if $|P_i| = 1$ no clusters exist, and if $|P_i| = 2$
at most two clusters exist: one for all genes where the solutions agree and one for where they disagree.

Given the useful set of clusters, we can now define how they are used.  P3 uses
crossover to recombine known information in a level of the pyramid with a candidate
solution.  For each cluster, do the following.  Select a random donor from the population.
Copy the donor's gene values for all loci in the cluster into the candidate solution.
Evaluate the candidate.  If the fitness of the candidate has been reduced, revert the candidate's
gene values.  To clarify, changes are kept in the case when the fitness is unaffected.
This allows the crossover to drift through fitness plateaus.

When choosing the donor P3 ensures there is at least one gene value
different between the donor and the candidate solution.  This can potentially require
up to $O(|C_j||P_i|)$, but empirically appears more efficient than allowing donations that
do not change any gene values.

To complete a single crossover event, all of the useful clusters are tested, each
requiring at most a single evaluation.  Clusters are tested in smallest first order,
with clusters of equal size ordered randomly.  This ensures smaller building blocks
are optimized before large changes are made to the genome, which helps preserve
diversity.  This also helps prevent genetic hitchhikers, which make interfere with
entropy calculations.  As cluster building and application ordering are not dependent on
the position of loci in the genome, there is no positional bias in the search points
evaluated.

@@

Similarly, by starting from a random position for each new solution, P3 progressively reduces
the chance of spurious linkage formation due to initialization.  Furthermore, these random
restarts provide exploration to balance out the hill climber and crossover's extreme
exploitation.  This also allows the population to grow large enough to contain enough
diversity to solve the problem without having to know the optimal size beforehand.

@@

\begin{comment}
A major advantage to the design of P3 is that it makes very weak assumptions about
the search landscape.  In many evolutionary systems the crossover operator assumes
that genes linked in the search landscape are collocated in the genome.  This means
a problem changes difficulty depending on how the genome is set up, meaning the user
has to know information about gene linkage when applying the algorithm.  All of P3's
operations ignore gene ordering, removing that level of configuration from the user.
\end{comment}

\subsection{Contrasted With The Generational Model}
A major advantage of P3's structure is that it avoids the need to set a fixed
population size.  In a generational model, the diversity contained in the initial
population represents a bottleneck for future generations.  This is true
because in most systems, especially on deceptive problems, mutation rates are
low enough as to make the creation of large, multi-gene substructures unlikely
after initialization.  As such the generational model often results in a race between
decreasing diversity and increasing fitness, with the hope that the global optimum
will be found before too much diversity is lost.  As such, the population size
must be set large enough to contain sufficient initial diversity and
to maintain that diversity during search for just long enough for all of it to be
leveraged.

Premature convergence of a population leads to the generational model often having
two termination conditions: out of time or out of diversity.  In the former, if
a smaller population size had been used, more of the diversity could have been exploited
meaning a better solution likely could have been reached in the same number of evaluations.
In the latter, if a larger population size had been used more evaluations could have been
performed increasing the likelihood of improving solution quality.  Often the only option
is to restart, throwing away all of the previous search information.  In either case, unless
you can set the population size correctly beforehand, the generational model will likely
result in waste.

As P3 has exploited all known information each time the top of the pyramid is reached,
there is no fear of wasted evaluations on unused diversity.  Similarly, because P3
does not throw away previously optimized solutions while adding diversity, it avoids
waste caused by premature convergence.

In general, P3 only makes two assumptions: local structure and non uniform gene interdependence.
In order for P3 to be efficient there needs to be at least some correlation between a
solution's fitness and the fitness of its single bit neighborhood.  This is required for
the hill climber to function properly.  Under this assumption, each time the hill climber
improves a solution it also improves the expected fitness of the neighborhood of the solution.
Stated another way, this assumes that small changes in genotype result in small changes in phenotype.
The non uniform gene interdependence assumption is required by the crossover operator to
perform clustering.  Without this assumption, the clusters formed by crossover will
all be due to spurious relations, and the donation of genes between solutions will be
no better than random modification.  Both assumptions are likely to hold true for
real world problems.  Also, many existing generational model systems have similar
or even more restrictive assumptions about search landscape.

\subsection{Order of Complexity}
With the exception of crossover, P3 requires no more than $O(N)$ computation
per evaluation, implying it is limited by the time required to evaluate a solution.
In order to perform crossover, P3 must spend $O(N^2)$ time updating pairwise
frequencies each time a solution is added to a level of the pyramid.  Similarly,
it requires $O(N^2)$ time to rebuild the set of crossover clusters once the frequency
table has been updated.  Na\"ively, this leads to the algorithm itself requiring $O(N^2)$ time.
However, multiple evaluations can occur for each time a single individual is added
to a population.

Each time a random restart occurs, at minimum $N+1$ evaluations are performed.
This means each time a solution is added to $P_0$, at least $O(N)$ evaluations
must be performed.  This implies that the $N^2$ cost of the update to $P_0$ can
be amortized over these evaluations, making this update $O(N)$ per evaluation.  Note
that $N+1$ is a very weak lower bound, as this only occurs if the randomly generated
individual is already at a local optimum.  Furthermore, only restarts that result
in a unique individual cause updates to $P_0$.

To cause an update to $P_i$, all of the crossovers discovered for $P_{i-1}$ must
first be applied.  Each causes their own evaluation.  If P3 were to na\"ively use
all crossovers and perform an evaluation each time, this would cause $O(N)$ evaluations.
Again, this would result in $O(N)$ amortized cost per update to $P_i$.  However,
Section~\ref{sec-crossover} discussed how P3 ignore some clusters and will skip
an evaluation if no change to the solution was made by a crossover.  As such we currently
are unable to state conclusively that updating $P_i$ will require $O(N)$ evaluations.  That
said, empirically we expect it to be so.  Furthermore, it seems counterintuitive that
intentionally wasting evaluations would be considered beneficial to runtime.

There is one more operation in P3 that may not scale linearly with genome size per evaluation:
finding a non-identical donor.  When performing crossover P3 iteratively searches a population
for a random donor which contains genes which differ from the current solution for the current crossover
cluster.  In the worst case, this operation could require $|C_j||P_i|$, where $|C_j|$ is the number
of genes in the crossover and $|P_i|$ is the number of solutions in the population.  However, $|P_i|$ grows
very slowly.  Also, in empirical evaluation, P3's wall clock time per evaluation appears to scale linearly with
problem size.

TODO Consider discussing memory usage.

@@

If P3's quality relies solely on starting at random points and applying hill climber,
this comparison algorithm should perform even better than P3.  Similar to P3 this
algorithm is Parameter-less, makes no assumptions about gene ordering, and will
find the global optima given a sufficient (potentially exponential) number of evaluations.

@@

There have been other model based optimization techniques that use little to
no parameters.  Specifically LO-LIMD~\cite{posik:2011:parameterless} uses a
population size of one and a parameter free algorithm to perform optimization.
It relies on the existence of completely separable subproblems and noise free evaluation in order to determine
if two genes are linked through perturbation of their values.  While effective on
this class of problems, most real world problems do not fall into this category.
In general, LTGA is the most powerful general optimizer with few parameters.

@@

In choosing test problems, we decided to use only benchmarks with knowable global
optima.  This allows us to measure how many evaluations each method requires to converge to the
global optima, as opposed to just the highest fitness reached.  This is advantageous
as three of the four algorithms have no hard termination condition.  The remaining
algorithm, LTGA, uses a population parameter which can be increased to slow
convergence with the potential increase in fitness reachable before convergence.
Therefore it is difficult to say how and when to declare any of the techniques
converged unless the global optimum is found.

In practice it is often more important to find a good solution quickly than find
the best solution after an exceedingly high number of evaluations.  However, even
a cursory consideration of P3's operators should convince you of its ability to
find high quality solutions in very few evaluations.  This is specifically true of the iterative
application of the first improvement hill climber and the exceptionally elitist
crossover operator.  Therefore we feel it is more important to test P3's resistance
to premature convergence, as that seems a more likely flaw than being slow to reach
``good enough'' solutions.

@@

The classic example for this kind of test problem is the Deceptive Trap problem.
For this benchmark the genome is broken up into non-overlapping problems of $k$
genes, with the fitness of a single ``trap'' given by Equation~\ref{eq-deceptivetrap}.
\begin{equation}
   trap(t) = \left\{
     \begin{array}{rl}
       k-1-t, &  t<k\\
       k,   &  t = k
     \end{array}
   \right.
  \label{eq-deceptivetrap}
\end{equation}
In this equation $t$ represents the number of bits set in the trap.  For each trap
there is one local optima (all bits set to 0) and one global optima (all bits set to 1).
We chose to use a trap size of $k=7$.
This problem requires exponential time to solve unless there is some method to effectively
move complete traps between solutions, making this an excellent crossover test problem.

However, when using a hill climber and linkage learning, the Deceptive Trap problem
may be trivially solved~\cite{goldman:2012:ltga}.  As such the Deceptive Step Trap
problem was defined, given in Equation~\ref{eq-deceptivesteptrap}.
\begin{equation}
   step\_trap(t) = \left \lfloor \frac{(k-s)\pmod{s} + trap(t)}{s} \right \rfloor
  \label{eq-deceptivesteptrap}
\end{equation}
This problem leverages the original Deceptive Trap problem definition, with the
addition of fitness plateaus of configurable size $s$.  This creates an enormous
number of local optima and makes detection of linkages much more challenging.  We
chose to use a trap size of $k=7$ with plateaus of size $s=2$.

@@

NK landscapes define a class of problems such that the fitness of a single gene's
value is dependent on $k$ other genes in the genome.  Finding the solution which
maximized fitness on an arbitrary NK landscape requires exponential time, and
are therefore a good candidate for heuristic techniques.  Nearest Neighbor NK
landscapes are a subset of NK landscapes such that each gene is dependent on the $k$
genes following it in the genome.  The advantage of this subset is that they can be
solved in polynomial time~\cite{wright:2000:solvingnk}.  For our experiments we
used wrapping neighborhoods, such that genes near the end of the genome depend on
genes at the start of the genome.  We chose to set $k=5$.

@@

Our final problem class is a subset of the Maximum Satisfiability (MAXSAT) problem.
This problem consists of a series of disjunctive clauses, such that each clause contains
three problem variables, some of which are negated. While
MAXSAT requires exponential time to solve, we have devised a method of generating
instances that ensures each has at least one solution that satisfies all clauses.
First, a random target solution is generated.  For each clause, set the sign of each
variable in the clause such that that clause will be satisfied by the target solution.
To ensure there is no bias to variable signs in the generated problem there is a $\frac{1}{6}$ probability
all clause signs match the target, a $\frac{1}{6}$ probability two of the three signs match,
and a $\frac{4}{6}$ probability only one sign matches.  By setting the clause signs
in this way we ensure the target solution is always able to satisfy all clauses,
meaning all problems have a known best fitness.  Note that its possible there to be
other solutions that also satisfy all clauses.  We chose a clause to variable ratio
of $r=4.27$.

\section{Experimental Results}
\begin{figure*}
  \centering
  \includegraphicswide{WithRast}
  \caption{Comparison of the median number of evaluations to reach the global optimum for
           the four different optimization methods on seven benchmark problems with respect
           to problem size.  If the median run did not reach the global optimum no point
           is drawn.  Results given on a log-log scale.}
  \label{fig-results}
\end{figure*}

@@

  \end{tabular}
	\caption{Comparison of the median evaluations to success for LTGA and P3.  Includes
        	the 95\% bootstrapped confidence interval around the median, as well as
        	the speedup factor for P3 over LTGA.}
	\label{table-results}
\end{table}

In total, our experiments performed 100 runs of each optimization algorithm across
7 benchmark problems, with each run limited to 100 million evaluations.
In performing data collection, we performed over 150,000 runs, totaling approximately
1.5 trillion evaluations.  Of these, approximately 100,000 runs were used to optimally
set the population size of LTGA.  Note that these runs should be considered a training phase,
as such all reported results for LTGA use independent runs to create a testing phase.
This was the only algorithm tested that received such tuning.  As such, LTGA should have a significant
advantage over the others in that it has pre-learned the correct population size.
All of the code used to perform our experiments
as well as individual run results and analysis scripts are available from our
website\footnote{\url{https://name.of.site.removed/anonymous}}.  The results
are summarized in Figure~\ref{fig-results} and Table~\ref{table-results}.

On all 7 problems, the First Improvement Hill Climber and the $(1+(\lambda,\lambda))$
optimizers appear to have a higher growth complexity than P3.  On all of the problems
except MAT-SAT they are also worse than LTGA.  This makes sense as all of the
benchmarks are deceptive in some sense, and neither of these techniques is designed
to handle deceptive landscapes.  The inability for the iterative hill climber to solve
any of the problems indicates that P3 must be doing something more than just local search.

Figure~\ref{fig-results} does provide one misleading
exception for $(1+(\lambda,\lambda))$ on the deceptive trap problem, where it appears
level out at high genome sizes.  This is entirely an artifact of the extremely high
variance of this algorithm on this problem.  Note that many of the large problem
sizes failed to find the optimum for the majority of runs (no dot drawn for that size).

The behavior of P3 and LTGA on the four single instance problems appears very similar,
with P3 generally being lower by a constant factor.  This makes sense as LTGA's
tuning allows it to determine a single population size sufficient to solve that
instance.  However, because it uses a fixed population size, it must set the that
size large enough to ensure all random initializations of that population are
likely to contain enough diversity to reach the global optimum.  By comparison, P3 is able to
iteratively increase the population size, stopping once enough diversity has been generated.
As such we would expect P3 to do better, but not a lot better.

When applied to the randomly generated problem classes, P3 appears to outperform LTGA
by an order of complexity.  We suspect some of the cause is LTGA's need to fix
a single population size.  Specifically, during tuning LTGA must set its population
size large enough to solve all of the random instances of that class.  Therefore
all instances which could be solved more efficiently with a smaller population size
are penalized.  P3 on the other hand grows with problem difficulty.  As such its
median evaluation time better reflects the median difficult of a class, whereas LTGA's
evaluation time reflects the maximum difficulty of a class.

Of the runs where LTGA was successful in the median run, it failed to find the optimal solution
109 times, or 0.8\%.  By comparison, P3 failed to find the optimum 3 times.  All three
where on Nearest Neighbor NK, two at size 550 and one at size 600.  Also, P3 was able
to reliably find the optimum on Nearest Neighbor NK $N=650$, $N=700$ and MAX-SAT $N=65$, $N=70$
while LTGA was not.

Table~\ref{table-results} provides a close up look at the largest problem sizes
in which LTGA was able to find the global optimum reliably in less than 100 million
evaluations.  For each, we performed a pairwise Mann-Whitney~U test, all resulting
in p-values less than $10^-22$, with the exception of Deceptive Step Trap which
resulted in $0.00025$.  As such these results are highly significant.  The amount
of speedup ranged from 1.1 on Deceptive Step Trap to 942 on MAX-SAT, but in all
cases P3 outperformed LTGA and all other comparison techniques.

As a further verification that our tuning for P3 is valid, we refer
to~\cite{thierens:2013:ltgahiff}, which tested the same variant of LTGA that we
used.  In the section on Child Node filtering, they reported approximately 200,000
mean evaluations to success on the HIFF problem using $N=512$.  Our result of
239,365 is of comparable quality, with P3 solving the same problem using an average
of 92,129 evaluations.  The discrepancy is likely due to the cited work using
their tuning runs in their average, unrealistically allowing each run to use a different tuned population size.

\section{Promising Theory}
Due to P3's persistent integration of random solutions, it always has a non zero
probability of reaching the global optimum.  Because it meets this prerequisite
(unlike LTGA), it is likely possible to derive expected running times for P3 on
a number of problems.  While we do not present a rigorous solution to any particular
problem here, we wish to illuminate potential avenues for future proofs.

The traditional One Max class, as well as all functions of unitation, will be
trivially solved by the hill climber in at most $N+1$ evaluations.  These come
from the initial evaluation of the random solution, plus the $N$ evaluations required
to test each gene for improvement.  While more difficult to bound, any unimodal
search space without fitness plateaus will be solved by a single application
of the hill climber.  Notably we expect a bound on Leading Ones of $O(N^2)$ time.
%TODO For future reference, n / factorial(n) + sum(i**2 / factorial(i + 1) for i in range(n))
%is the expected number of improvements made per iteration (converges to 1.718281828459045).
